---
title: "Worldwide Health Outcomes"
subtitle: "Explaining variation in health measures around the world"
author: "the-gr8est-team <br> Roni O., Judy Z, Matthew W."
institute: "Duke University"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      ratio: "16:9"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      

---

```{r load-packages, include = FALSE}
# Add any additional packages you need to this chunk
# Remove any packages from this list that you're not using
library(tidyverse)
library(tidymodels)
library(palmerpenguins)
library(knitr)
library(xaringanthemer)
library(infer)
library(ggplot2)
library(sf)
#install.packages("rnaturalearth")
library(rnaturalearth)
#install.packages("rnaturalearthdata")
library(rnaturalearthdata)
library(rgeos)
library(scales)
#install.packages("grid")
library(grid)
library(gridExtra)
#install.packages("kableExtra")
library(kableExtra)
```

```{r setup, include=FALSE}
# For better figure resolution
knitr::opts_chunk$set(
  fig.retina = 3, 
  dpi = 300, 
  fig.width = 6, 
  fig.asp = 0.618, 
  out.width = "77%"
  )
```

```{r load-data, include=FALSE}
# Load your data here
full_data <- read_csv("Full Dataset.csv")
```

```{r style-slides, echo = FALSE}
style_xaringan(
  title_slide_background_image = "img/watercolour_sys02_img32_percolate.jpg"
)
```


## Background

--

### - GDP, Total Health Expenditure (THE), and Education shown to positively correlate with health outcomes

--

### - Life Expectancy = projected life span from birth

--

### - Under 5 Mortality Rate = predicted number of deaths before age 5 (per 1000)

---

class: inverse, middle, center

## Research Question

# This analysis aims to determine the significance of relationships between a set of World Development Indicators and health outcomes, measured by life expectancy and under-5 mortality rate.


---

# Data Overview

--

### - Health Spending = from Global Health Data Exchange

--

### - Economic/Health-Related Data = World Bank World Development Indicators

---

## Summary Stats - Response Variable
```{r outcome-histo, warning = FALSE, echo=FALSE, fig.align = "center"}
p_lifeexp <- full_data %>%
  ggplot(aes(full_data$'Total Life Expectancy')) + 
  geom_histogram(fill = "blue", na.rm = TRUE, bins = 50) + 
  labs(
    x = "Life Expectancy (Years)",
    y = "Frequency",
    title = "Distribution of Life Expectancy"
  ) 

p_fert <- full_data %>%
  ggplot(aes(full_data$'Fertility rate, total (births per woman)')) + 
  geom_histogram(fill = "blue", na.rm = TRUE, bins = 50) + 
  labs(
    x = "Fertility Rate (Births per Woman)",
    y = "Frequency",
    title = "Distribution of Fertility Rate"
  ) 

p_log_fert <- full_data %>%
  ggplot(aes((log(full_data$'Fertility rate, total (births per woman)')))) + 
  geom_histogram(fill = "blue", na.rm = TRUE, bins = 50) + 
  labs(
    x = "Fertility Rate (Log of Births per Woman)",
    y = "Frequency",
    title = "Fertility Rate"
  )

p_mort <- full_data %>%
  ggplot(aes(full_data$`Mortality rate, under-5 (per 1,000 live births)`)) + 
  geom_histogram(fill = "blue", na.rm = TRUE, bins = 50) + 
  labs(
    x = "Mortality Rate (per 1,000 Live Births)",
    y = "Frequency",
    title = "Mortality Rate"
  )

p_log_mort <- full_data %>%
  ggplot(aes(log(full_data$`Mortality rate, under-5 (per 1,000 live births)`))) + 
  geom_histogram(fill = "aquamarine", na.rm = TRUE, bins = 50) + 
  labs(
    x = "Mortality Rate (Log per 1,000 Live Births)",
    y = "Frequency",
    title = "Distribution of Mortality Rate"
  ) + theme(axis.title.x = element_text(size = 10))
grid.arrange(grobs = list(p_lifeexp, p_log_mort), ncol = 2)
```

---
## Summary Stats - Predictors

```{r healtcarespendinghisto, warning = FALSE, echo=FALSE, fig.align = "center"}
p_the <- full_data %>%
  ggplot(aes(x = the_per_cap_ppp_mean)) + 
  geom_histogram(fill = "red", na.rm = TRUE, bins = 30) + 
  labs(
    x = "Mean Total Health Spending per Person (in PPP dollars)",
    y = "Frequency",
    title = "Distribution of Health Care Spending"
  )

p_log_the <- full_data %>%
  ggplot(aes(x = log(the_per_cap_ppp_mean))) + 
  geom_histogram(fill = "red", na.rm = TRUE, bins = 30) + 
  labs(
    x = "Log Mean THE per Capita (Log PPP dollars)",
    y = "Frequency",
    title = "Health Care Spending"
  ) +
  theme(title = element_text(size = 8),
        axis.title.x = element_text(size = 5),
        axis.title.y = element_text(size = 8))

p_log_pov <- full_data %>%
  ggplot(aes(log(full_data$`Poverty headcount ratio at national poverty lines (% of population)`))) + 
  geom_histogram(fill = "blue", na.rm = TRUE, bins = 30) + 
  labs(
    x = "Log Poverty Headcount Ratio",
    y = "Frequency",
    title = "Poverty Headcount"
  )

p_pov <- full_data %>%
  ggplot(aes(full_data$`Poverty headcount ratio at national poverty lines (% of population)`)) + 
  geom_histogram(fill = "maroon1", na.rm = TRUE, bins = 30) + 
  labs(
    x = "Poverty Headcount Ratio",
    y = "Frequency",
    title = "Poverty Headcount"
  ) +
 theme(title = element_text(size = 8),
        axis.title.x = element_text(size = 5),
        axis.title.y = element_text(size = 8))
p_log_gdp <- full_data %>%
  ggplot(aes(log(full_data$`GDP per capita (current US$)`))) + 
  geom_histogram(fill = "darkorange", na.rm = TRUE, bins = 30) + 
  labs(
    x = "Log GDP",
    y = "Frequency",
    title = "GDP"
  ) +
 theme(title = element_text(size = 8),
        axis.title.x = element_text(size = 5),
        axis.title.y = element_text(size = 8))

p_gdpg <- full_data %>%
  ggplot(aes(log(full_data$`GDP per capita growth (annual %)`))) + 
  geom_histogram(fill = "darkgoldenrod2", na.rm = TRUE, bins = 30) + 
  labs(
    x = "GDP Growth (Annual %)",
    y = "Frequency",
    title = "GDP Growth"
  ) +
 theme(title = element_text(size = 8),
        axis.title.x = element_text(size = 5),
        axis.title.y = element_text(size = 8))

p_inc <- full_data %>%
  ggplot(aes(full_data$`Income share held by lowest 20%`)) + 
  geom_histogram(fill = "purple", na.rm = TRUE, bins = 30) + 
  labs(
    x = "Income Share Held by Lowest 20%",
    y = "Frequency",
    title = "Income Share (%)"
  ) +
   theme(title = element_text(size = 8),
        axis.title.x = element_text(size = 5),
        axis.title.y = element_text(size = 8))

p_edu <- ggplot(data = full_data, aes(full_data$`Educational attainment, at least completed upper secondary, population 25+, total (%) (cumulative)`)) + 
  geom_histogram(fill = "yellow", na.rm = TRUE, bins = 30) + 
  labs(
    x = "Secondary Education Completion (%)",
    y = "Frequency",
    title = "Educational Attainment"
  ) +
   theme(title = element_text(size = 8),
        axis.title.x = element_text(size = 5),
        axis.title.y = element_text(size = 8))

grid.arrange(grobs = list(p_log_the, p_log_gdp, p_gdpg, p_pov, p_inc, p_edu), ncol = 3)
```

---

## Summary Stats - MAPS!

```{r healthspending_and_lifeexpectancymap, warning =FALSE, echo=FALSE, fig.align = "center"}

world <- ne_countries(scale = "medium", returnclass = "sf")
full_data1 = full_data %>%
  filter(Year == 2015)
healthspending.world <- merge(world, full_data1, by.x="adm0_a3", by.y="Country Code")

map1 <- ggplot(data = healthspending.world) +
    geom_sf(aes(fill = log(the_per_cap_ppp_mean))) +
    scale_fill_viridis_c(option = "plasma") +
    labs(title = "Health Spending",
       fill = "Log(THE) (log PPP dollars)") +
    theme(legend.title = element_text(angle = 90, size = 8),
        legend.key.width = unit(2, 'mm'),
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) + 
    guides(fill = guide_colorbar(title.position = "left"))


map2 <- ggplot(data = healthspending.world) +
    geom_sf(aes(fill = log(`Mortality rate, under-5 (per 1,000 live births)`))) +
    scale_fill_viridis_c(option = "plasma") +
    labs(title = "Mortality Rates",
       fill = "Log(Mortality Rates Under 5, per 1000)")  +
    theme(legend.title = element_text(angle = 90, size = 7),
        legend.key.width = unit(2, 'mm'),
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) + 
    guides(fill = guide_colorbar(title.position = "left"))

map3 <- ggplot(data = healthspending.world) +
    geom_sf(aes(fill = `Total Life Expectancy`)) +
    scale_fill_viridis_c(option = "plasma") +
    labs(title = "Life Expectancies",
       fill = "Life Expectancy (Years)")  +
    theme(legend.title = element_text(angle = 90, size = 8),
        legend.key.width = unit(2, 'mm'),
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) + 
    guides(fill = guide_colorbar(title.position = "left"))

grid.arrange(grobs = list(map2, map3), ncol = 1, top = "Geographic Trends of Health Measures and Health Expenditure")
```
---

## Modelling - Life Expectancy

$\text{Total Life Expectancy} = \\\beta_0 + \beta_1*log(\text{THE}) + \beta_2*log(\text{THE})^2 + \\\beta_3*log(\text{GDP}) + \beta_4*\text{Poverty} + \beta_5*\text{Education} + \\\beta_6*\text{Education}^2 + \beta_7*\text{GDP Growth} + \beta_8*\text{Poverty}*\text{Education} + \\\beta_9*\text{GDP Growth}*\text{Education} + \beta_{10}*\text{Poverty}*log(\text{GDP}) + \\\beta_{11}*log(\text{GDP})*\text{Education} + \beta_{12}*log(\text{THE})*\text{Poverty}$ 

```{r mixed_life_expectancy_model_table, include = FALSE}
mixed_data_life_exp <- full_data %>%
  filter(is.na(`Total Life Expectancy`) == FALSE) %>%
  filter(`Total Life Expectancy` != '..') %>%
  filter(is.na(the_per_cap_ppp_mean) == FALSE) %>%
  filter(is.na(`GDP (current US$)`) == FALSE) %>%
  filter(is.na(`GDP growth (annual %)`) == FALSE) %>%
  filter(is.na(`Poverty headcount ratio at national poverty lines (% of population)`) == FALSE) %>%
  filter(is.na(`Educational attainment, at least completed upper secondary, population 25+, total (%) (cumulative)`) == FALSE) %>%
  
  mutate(spending = log(the_per_cap_ppp_mean)) %>%
  mutate(gdp = log(`GDP (current US$)`)) %>%
  mutate(gdpg = `GDP growth (annual %)`) %>%
  mutate(poverty = `Poverty headcount ratio at national poverty lines (% of population)`) %>%
  mutate(education = `Educational attainment, at least completed upper secondary, population 25+, total (%) (cumulative)`)

mixed_data_life_exp$spending2 <- mixed_data_life_exp$spending^2
mixed_data_life_exp$education2 <- mixed_data_life_exp$education^2

lifeexp_mixedfit<- linear_reg() %>%
  set_engine("lm") %>%
  fit(`Total Life Expectancy` ~  
        spending 
      + spending2 
      + gdp
      + poverty
      + education 
      + education2
      + poverty*education
      + gdpg*education
      + gdp*poverty
      + gdp*education
      + spending*poverty
      ,data = mixed_data_life_exp)

tidy(lifeexp_mixedfit, conf.int=TRUE)
```

``` {r table3, echo = FALSE, fig.align = "center"}
result3 <- tidy(lifeexp_mixedfit, conf.int=TRUE)
vals3 <- c("$\\beta_0$", "$\\beta_1$", "$\\beta_2$", "$\\beta_3$", "$\\beta_4$", "$\\beta_5$", "$\\beta_6$", "$\\beta_7$", "$\\beta_8$", "$\\beta_9$", "$\\beta_{10}$", "$\\beta_{11}$", "$\\beta_{12}$")
result3 <- result3[-c(1, 3:4)]
result3 <- t(result3)
colnames(result3) <- vals3
rownames(result3) <- c("Value", "p-value", "CI Lower", "CI Higher")
result3 %>% kable(escape = FALSE,
                  digits = 4) %>%
  kable_classic(full_width = F, html_font = "helvetica") %>%
  kable_styling(latex_options = c("hold_position","scale_down"),
                font_size = 16)
```

---

## Modelling - Life Expectancy (Residuals)

```{r mixed_life_expectancy_model_resid, warning = FALSE, echo = FALSE, fig.height = 3}
lifeexp_mixedfit_aug <- augment(lifeexp_mixedfit$fit)

ggplot(lifeexp_mixedfit_aug, mapping = aes(x = .fitted, y = .std.resid)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, color = "blue", lty = "dashed") +
  labs(title = "Quadratic Interaction Model",
       x = NULL,
       y = NULL)  +
  theme(title = element_text(size = 8))
```

---

## Modelling - Mortality Rates Under 5

$\log(\text{Under-5 Mortality}) = \\\beta_0 + \beta_1*log(\text{THE}) + \beta_2*log(\text{THE})^2 + \beta_3*log(\text{GDP}) + \beta_4*log(\text{GDP})^2 + \\\beta_5*\text{Income} + \beta_6*\text{Income}^2 + \beta_7*\text{Poverty}^2 + \beta_8*\text{Education} + \\\beta_9*\text{Education}^2 + \beta_{10}*\text{GDP Growth} + \beta_{11}*\text{Poverty} + \beta_{12}*\text{Income}*\text{Education} + \\\beta_{13}*log(\text{THE})*log(\text{GDP}) + \beta_{14}*\text{Income}*\text{GDP Growth} + \\\beta_{15}*\text{Education}*\text{GDP Growth} + \beta_{16}*log(\text{GDP})*\text{Education} + \\\beta_{17}*log(\text{GDP})*\text{Income} + \beta_{18}*\text{Income}*\text{Poverty}$ 

```{r mixed_mortality_model_table, include = FALSE}
mixed_data_mortality <- full_data %>%
  filter(is.na(`Mortality rate, under-5 (per 1,000 live births)`) == FALSE) %>%
  filter(is.na(the_per_cap_ppp_mean) == FALSE) %>%
  filter(is.na(`GDP (current US$)`) == FALSE) %>%
  filter(is.na(`GDP growth (annual %)`) == FALSE) %>%
  filter(is.na(`Income share held by lowest 20%`) == FALSE) %>%
  filter(is.na(`Poverty headcount ratio at national poverty lines (% of population)`) == FALSE) %>%
  filter(is.na(`Educational attainment, at least completed upper secondary, population 25+, total (%) (cumulative)`) == FALSE) %>%
  
  mutate(mortality = log(`Mortality rate, under-5 (per 1,000 live births)`)) %>%
  mutate(spending = log(the_per_cap_ppp_mean)) %>%
  mutate(gdp = log(`GDP (current US$)`)) %>%
  mutate(gdpg = `GDP (current US$)`) %>%
  mutate(income = `Income share held by lowest 20%`) %>%
  mutate(poverty = `Poverty headcount ratio at national poverty lines (% of population)`) %>%
  mutate(education = `Educational attainment, at least completed upper secondary, population 25+, total (%) (cumulative)`)

mixed_data_mortality$spending2 <- mixed_data_mortality$spending^2
mixed_data_mortality$gdp2 <- mixed_data_mortality$gdp^2
mixed_data_mortality$income2 <- mixed_data_mortality$income^2
mixed_data_mortality$poverty2 <- mixed_data_mortality$poverty^2
mixed_data_mortality$education2 <- mixed_data_mortality$education^2

mortality_mixedfit <- linear_reg() %>%
  set_engine("lm") %>%
  fit(mortality ~  spending 
      + spending2
      + gdp
      + gdp2
      + income
      + income2
      + poverty2
      + education
      + education2
      + income*education
      + spending*gdp
      + gdpg*income
      + gdpg*education
      + gdp*education
      + gdp*income
      + income*poverty
      ,data = mixed_data_mortality)

tidy(mortality_mixedfit, conf.int=TRUE)
```

``` {r table6, echo = FALSE, fig.align = "center"}
result6 <- tidy(mortality_mixedfit, conf.int=TRUE)
vals6 <- c("$\\beta_0$", "$\\beta_1$", "$\\beta_2$", "$\\beta_3$", "$\\beta_4$", "$\\beta_5$", "$\\beta_6$", "$\\beta_7$", "$\\beta_8$", "$\\beta_9$", "$\\beta_{10}$", "$\\beta_{11}$", "$\\beta_{12}$", "$\\beta_{13}$", "$\\beta_{14}$", "$\\beta_{15}$", "$\\beta_{16}$", "$\\beta_{17}$", "$\\beta_{18}$")
result6 <- result6[-c(1, 3:4)]
result6 <- t(result6)
colnames(result6) <- vals6
rownames(result6) <- c("Value", "p-value", "CI Lower", "CI Higher")
knitr::kable(result6,
             escape = FALSE,
             digits = 3) %>%
  kable_styling(latex_options = c("hold_position","scale_down"),
                font_size = 12) %>%
  kable_classic(html_font = "helvetica")
```

---

## Modelling - Mortality Rates Under 5 (Residual)

```{r mixed_mortality_model_resid, warning = FALSE, echo = FALSE}
mortality_mixedfit_aug <- augment(mortality_mixedfit$fit)

ggplot(mortality_mixedfit_aug, mapping = aes(x = .fitted, y = .std.resid)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, color = "blue", lty = "dashed") +
  labs(title = "Quadratic Interaction Model",
       x = NULL,
       y = NULL)  +
  theme(title = element_text(size = 8))
```

---

# Conclusions

--

### - Education!!!

--

### - Other findings inconsistent...

---

# Limitations

### - Limited data from low-income countries

--

### - Overfitting

--

### - Some coefficients difficult to interpret

--

### - Some clustering in residual plot

--

### - No statements about causality
